```{r, echo=F, results='hide', message =F, warning=F}
# Check you have them and load them
list.of.packages <- c("kableExtra", "tidyr")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
lapply(list.of.packages, require, character.only = TRUE)


indDat  <- read.csv("data/processed_data/Algar_30min_Independent.csv", header=T)

dLookup <- read.csv("data/processed_data/Algar_daily_effort_lookup.csv", header=T) 

monCount <- read.csv("data/processed_data/Algar_30min_Independent_Monthly_counts.csv", header=T)
```

# Common analysis data formats

Although the types of analysis you can perform on camera trap data vary markedly, they often depend on three key dataframe structures. We introduce these structures here, then show you how to apply them in subsequent chapters. 

## Independent detections
The independent detections dataframe is the work horse of all camera trap analyses, it is from this that you build the rest of your data frames. The threshold we use for determining what is an "independent detection" is typically 30 minutes... because camera trappers are creatures of habit! If you want to dig a little deeper it to the why, there is a nice summary in [Rahel Sollmans "A gentle introduction to camera‐trap data analysis"](https://onlinelibrary.wiley.com/doi/abs/10.1111/aje.12557):

*Researchers have used different thresholds, typically 30 min (e.g., O'Brien, Kinnaird, & Wibisono, 2003) to an hour (Bahaa‐el‐din et al., 2016); some researchers have argued that multiple pictures within the same day may not represent independent detections (Royle, Nichols, Karanth, & Gopalaswamy, 2009). In most cases, this threshold is determined subjectively, based on the best available knowledge of the species under study. But it can also be determined based on the temporal autocorrelation (Kays & Parsons, 2014) or analysis of time intervals (Yasuda, 2004) of subsequent pictures.*

The default setting in our data processing script is 30 minutes, although it can be changed! 

**NOTE FOR THE FUTURE** add details on Event.ID	 Event.Duration	 Event.Groupsize	 Event.Observations columns

An independent dataframe should look something like this:

```{r, echo=F}
kbl(indDat)%>%
  kable_paper() %>%
  scroll_box(height = "200px")
```

## Effort look-up
Image data without effort data is worthless! There are lots of instances where you need to know which stations were operating on a given day. Some people like to store this information in a siteXdate matrix, but they are actually not that easy to data wrangle with. A long data frame with a site and date column is the most flexible (and keeps the dates in their native POSIX formats).

A typical day lookup looks like this:

```{r, echo=F}
kbl(dLookup)%>%
  kable_paper() %>%
  scroll_box(height = "200px")
```


## Counts/detections by time interval
We saved the most useful data format until last! A site, time interval, effort, and species  count dataframe integrates the independent data and daily lookup described above. You ca use it to create detection rates, occupancy data frames and much more!

We export yearly, monthly, weekly and daily data frames from our single site exploration script (which should cover you for much of what you want to do).

We include two different types of response terms:

  - Observations = the number of independent detections per time interval
  - Counts = sum of the independent minimum group sizes per time interval
  
Below is a typical count_month dataframe:

```{r, echo=F}
kbl(monCount)%>%
  kable_paper() %>%
  scroll_box(height = "200px")
```




