# Analysis covariates {#covariates}

Once we have created the building blocks for our data analysis dataframes, we must bring in the variables which will be used in the modelling steps. It is important to not that there are millions of ways to add covariates - both in terms of how you do it, and where you derive the data from. The covariates you use will depend on the questions you have, and the context of your survey. The examples provided here are not comprehensive and serve only as a guide!

**Create a new .R script**

Call it `02_example_covariates.R`.

**Load the required packages**

```{r ch4_1, echo=T, results='hide', message =F, warning=F, class.source="Rmain"}
library(kableExtra);library(dplyr); library(sf); library(modisfast); library(lubridate); library(corrplot); library(traitdata); library(terra); library(osmdata); library(elevatr)
```

We can simplify the covariate options we have available into two distinct categories:

**- Species traits**

The traits are species-level covariates which we think are important in structuring their responses to other covariates, such as human modification.

```{r ch4_2, class.source="Rmain"}
# Start by reading in your species list
sp_summary <- read.csv("data/processed_data/AlgarRestorationProject_species_list.csv", header=T)
```

**- Location-level covariates**

Location-level covariates are characteristics of the camera locations which are either fundamental to your question (such as the habitat type, degree of human modification, or distance to the nearest road), or they are things you are not directly interested in but must account for in your analyses. The way we derive and treat these variables are identical however.

```{r ch4_3, class.source="Rmain"}
locs <- read.csv("data/processed_data/AlgarRestorationProject_camera_locations.csv", header=T)
```

## Species traits

It is easier than ever before to add trait data to your species lists, particularly with the advent of R packages which pool multiple data sources such as the [`traitdata` database](https://github.com/RS-eco/traitdata), which to date, compiles data from 32 different sources.

Below we use this package to add trait data to the project species list:

```{r ch4_4, class.source="Rmain"}

# This package isn't available on Cran, so we must use the remotes package
#library(remotes)
#remotes::install_github("RS-eco/traitdata", build_vignettes = T, force=T)
# Load the library
library(traitdata)
```

To pull the data for a specific database we use the following code:

```{r ch4_5, class.source="Rmain"}
data("elton_mammals")
```

To explore the full list of available datasets click [this link](https://opentraits.org/datasets.html).

Let's take a look at what categories we have available to us:

```{r ch4_6, class.source="Rinfo", include=T}
 head(elton_mammals) %>% kbl() %>% scroll_box(height = "200px") %>%
  kable_paper("striped", full_width = F)
```

Lets make a new column `sp` which matches the species column in our 'sp_summary' dataset. We will use this as the "key" variable to extract the trait data.

```{r ch4_7, class.source="Rmain"}
elton_mammals$sp <- paste0(elton_mammals$Genus,"." ,elton_mammals$Species)
```

We do not want to take all of the trait data, so lets subset to `BodyMass.Value` and the activity data `Activity.Nocturnal  Activity.Crepuscular    Activity.Diurnal`.

```{r ch4_8, warning=F, message=F, class.source="Rmain"}
tmp <- elton_mammals[, c("sp","BodyMass.Value", "Activity.Nocturnal", "Activity.Crepuscular",	"Activity.Diurnal")]

# Lets rename the columns to make them more usable
tmp <- tmp %>% rename(
              mass_g = BodyMass.Value,
              act_noct = Activity.Nocturnal,
              act_crep = Activity.Crepuscular,
              act_diur = Activity.Diurnal)

sp_summary <- left_join(sp_summary, tmp)
```

And then check our output:

```{r ch4_9, class.source="Rinfo"}
sp_summary %>% kbl() %>% scroll_box(height = "200px") %>%
  kable_paper("striped", full_width = F)
```

If there are any NA's, it could be for several reasons:

i)  There is no trait data for that species - in this case you could either:

-   leave them as NA's (excluding them from later analyses) or if you are lucky, your analysis framework might be able to accommodate missing trait data

-   Give the species the mean values obtained from other species in its genus

ii) There is a mismatch in taxonomic resolution - you are working with a subspecies that isn't recognized. This is the case here with elk! Lets replace it with the data for (**Cervus elaphus**)

```{r ch4_9a, , class.source="Rmain"}

sp_summary[sp_summary$sp=="Cervus.canadensis", c("mass_g", "act_noct","act_crep","act_diur")] <- 
    elton_mammals[elton_mammals$sp=="Cervus.elaphus", c("BodyMass.Value", "Activity.Nocturnal", "Activity.Crepuscular",	"Activity.Diurnal")]
```

Whatever you do, remember to report it in your methods section!

Let's save our species list for a rainy day!

```{r ch4_10, class.source="Rmain"}
write.csv(sp_summary, paste0("data/processed_data/", locs$project_id[1],"_species_list.csv"), row.names = F)
```

## Camera station covariates

It is common to have a suite of covariates which you would like to investigate the effects of in your datasets. These could take the form of habitat designations or treatment types. These may already be included with your deployment data, or you may need to derive them from a variety of remote sources. In their simplest form, these variable are time invariant (they do not change), however you may have variables which change through time as well (we discuss these at the end). In the following steps, we walk through the process of manipulating and deriving example covariates.

For the time invariant covariates, we will add them to our `locs` dataframe imported above.

### Locally collected covariates

You may have collected some data in the field when deploying or checking your camera traps, and kept that data separate from your camera trap data (e.g. vegetation assessments). Provided that the naming convention you gave to these dataframes is the same as in your camera data (e.g. the location is in a column called `placename`) - you can do a 'left_join()\` to merge the two datasets.

Import a sample set of local covariates:

```{r ch4_11, class.source="Rmain"}
local_covs <- read.csv("data/raw_data/example_covariates/example_dataframe.csv")
```

Lets take a look at the data structure:

```{r ch4_12, echo=F, class.source="Rinfo"}
local_covs %>% kbl() %>% 
  kableExtra::scroll_box(height = "250px")
```

It is a dataframe where the survey locations are rows and the local covariates, in this case `line_of_sight_m`, are columns.

To add this data to our station data, we use a `left_join()` operation from the `dplyr()` package. It uses a key variable which is common in both data frames to add data from the "right-hand side" to the rows in the "left-hand side" which are not already present. Any rows present in the right-hand side which are not in the left-hand side will be skipped.

```{r ch4_13, message=F, warning=F, class.source="Rmain"}
locs <- left_join(locs, local_covs)   # From the dplyr package
```

For more examples of joins using `dplyr()` see: <https://dplyr.tidyverse.org/reference/mutate-joins.html>

### Remotely collected covariates

To exploit remotely collected data sources we need to use a package to help us with spatial data.

#### Key skills: `sf` package

The most intuitive package to learn spatial operations in R is the `simple features` package (a.k.a. `sf`). `sf` allows you to use spatial dataframes in the style of a typical R dataframe. We use this package frequently as it allows you to rapidly change coordinate projection systems (e.g. lat/long to UTM) and rapidly perform spatial operations.

Lets convert our "normal" dataframe to an `sf` dataframe:

```{r ch4_14, class.source="Rmain"}
locs_sf <- st_as_sf(locs,                              # We specify the dataframe 
                    coords=c("longitude", "latitude"), # The XY coordinates
                    crs=4326)                          # And the projection code
```

What does an `sf` object look like? Like a normal dataframe but with a weird header:

```{r  ch4_15, class.source="Rinfo"}
locs_sf
```

That header is important - it tells you the type of data you have (lines, points, polygons etc), and the projection information (`CRS`).

We like using `sf` as it is very easy to transform coordinates into different projections using `st_transform()`. Use the website [epsg.io](https://epsg.io/) to find the CRS codes for the projection you want - e.g. UTM 12N = 26712, then plug it into the following:

```{r ch4_16, eval=T, class.source="Rinfo"}
locs_utm <- st_transform(locs_sf, crs=26712)
```

Check the header of `locs_utm` and you should see that the CRS has changed!

Plotting `sf` objects is a little bit odd at first. If you try to plot them normally you get lots of replicated plots (one for each column) - try it:

```{r ch4_17, class.source="Rinfo"}
plot(locs_utm)
```

It can be useful as it varied the colors based on the properties of the column. Typically, however, we just want to plot the points themselves. We do that by wrapping the object in `st_geometry()` this just extracts the geometry of the object.

```{r ch4_18, class.source="Rinfo"}
plot(st_geometry(locs_utm))
axis(1)
axis(2)
```

We will use `st_geometry()` frequently below.

For more in depth information of `sf` functionality see: <https://r-spatial.github.io/sf/articles/sf1.html>

### Extracting data from local rasters

Often we have raster data layers stored which we would like to link to our camera locations. We have included one such example here, a raster which reflects the depth from the soil surface to the water table - a proxy for habitat type in this study site. The layer comes from the 1m Wet Area Mapping (WAM) layer:

[White, Barry, et al. "Using the cartographic depth-to-water index to locate small streams and associated wet areas across landscapes." Canadian Water Resources Journal 37.4 (2012): 333-347.](https://www.tandfonline.com/doi/full/10.4296/cwrj2011-909)

*NOTE* the raster has been down scaled to reduce its size for this course - it is no longer at 1m resolution.

The only time we deviate from the `sf` package is to deal with rasters. Raster objects in R are processed really slowly, especially if the raster is large. So instead we use the `terra` package.

```{r  ch4_19, class.source="Rmain"}
library(terra)
# Import the example raster using the stars package
ras <- rast("data/raw_data/example_covariates/example_raster.tif")
# Covert your sf locations to the same projection as your raster then put it in terra `vect` format,
locs_terra <- locs_sf %>% 
                st_transform(crs=st_crs(ras)) %>% # change the projection to match the raster
                vect() # Turn it into a terra object

```

Lets check our layers match up!

```{r ch4_20, class.source="Rinfo"}
plot(ras) # The terra package makes nice raster plots with legends
plot(locs_terra, add=T) # Add the survey locations as black dots
```

Great! Now lets buffer our camera locations by 250 meters, and take the average depth to water for each location:

```{r ch4_21, class.source="Rmain"}
# Buffer by 250m
locs_terra <- buffer(locs_terra,250)

# Extract the values to a temporary object - tmp 
tmp <- terra::extract(ras, locs_terra, fun=mean)

# Make a new column in locs_sf called water_depth_m
# They are ordered the same way so no need for a fancy join
locs_sf$water_depth_m <- tmp$Depth2WatAlgar

```

Finally, lets check the distribution of our data!

```{r ch4_22, class.source="Rinfo"}
# Basic boxplot in base R
boxplot(locs_sf$water_depth_m)
```

Most locations are on the water table (lowland sites), others are above it (upload sites), and they have different vegetation characteristics in the field.

### elevatr package

Camera studies often occur over a range of elevations - and we can quickly extract these elevations using the `elevatr` package and an `sf` dataframe.

There are two options, `epqs`:

```{r ch4_23, class.source="Rmain", warning=F, message=F}
library(elevatr)
locs_sf <- get_elev_point(locs_sf, 
                          src="epqs") # Can be slow if you have >500 camera points
```

```{r ch4_23b, class.source="Rmain", warning=F, eval=F}
locs_sf <- get_elev_point(locs_sf, 
                          src="aws", #Amazon Web Service Terrain Tiles - available globally
                          z = 12)  # z specifies the zoom level -> the lower the value the faster the code runs, but the coarser the elevation values are
```

The `src` option specifies the sources of the DEM data. We use `aws` Amazon Web Service Terrain Tiles - which are available globally.

The `z` option specifies the resolution of the underlying DEM, the high the value, the more detailed it is. However, it will take longer to run so do not go crazy.

Regardless of which method you use, let's plot the output:

```{r ch4_24, class.source="Rinfo"}
boxplot(locs_sf$elevation)
```

An elevation of \~ 500m was expected. Great!

If you want to download a full elevation raster for your area of interests, see the [`introduction to elevatr`](https://cran.r-project.org/web/packages/elevatr/vignettes/introduction_to_elevatr.html)

### Open Street Maps

Open Street Map (OSM) is an incredible resource for generating covariates for camera trap studies. For example, we might be interested in the distance to the nearest rivers, roads, or trails. All of these anthropogenic features are available in OSM!

**CAREFUL** OSM data is user contributed and often incomplete and patchy. Always plot your data and never assume it is complete without checking it first. For an example fo this see `water bodies` below.

First lets load the `osmdata` package.

```{r ch4_25, class.source="Rmain"}
library(osmdata)
```

The types of features we can extract using the `osmdata` package are listed here: <https://wiki.openstreetmap.org/wiki/Map_features>.

#### Highways

Camera trap projects are often interested in human disturbance, of which, highways are an important part.

Let's start by defining our area of interest. All `osmdata` queries begin with a bounding box defining the area of the query:

```{r ch4_26, class.source="Rmain"}
# First buffer our points by 10km to create an area of interest (aoi)
aoi <- st_bbox(st_buffer(locs_sf, 10000)) # Units are in meters 

```

We then use this bounding box to return all of the features which cross into it:

```{r ch4_27, class.source="Rmain"}
highway <- opq(aoi) %>% #using the bounding box
           add_osm_feature(key="highway") %>% #extract all highway features
           osmdata_sf()  # convert them into simple features format

```

The data you extract is its own "class" of data made up from multiple data types:

```{r ch4_28, eval=F, class.source="Rinfo"}
str(highway)
```

Which looks very intimidating! However, the key thing is that it is made up of multiple data slices, each of which represents an `sf` dataset. Let's take a look at three of these

-   \$osm_points
-   \$osm_lines
-   \$osm_polygons

```{r ch4_29, class.source="Rinfo"}
par(mfrow=c(1,3))
plot(st_geometry(highway$osm_points), main="osm_points")
plot(st_geometry(highway$osm_lines), main="osm_lines")
plot(st_geometry(highway$osm_polygons), main="osm_polygons")
```

The points or the lines datasets look must useful to us, there is nothing in the polygon layer.

Let's use the lines element and add out camera stations:

```{r ch4_30, class.source="Rinfo"}
par(mfrow=c(1,1))
plot(st_as_sfc(aoi))     # st_as_sfc created a polygon from a `bbox` object
plot(st_geometry(highway$osm_lines), add=T)
plot(st_geometry(locs_sf), col="red", add=T)
```

We can now calculate the distances from our cameras to these objects using the following codes:

`st_nearest_feature` gives us the index number of the feature which is closest to each station.

We can the use this to request the distance from that nearest feature to each camera station using `st_distance`. Which, put together, looks like:

```{r ch4_31, class.source="Rmain"}
# Create an index of the nearest object in `highway$osm_lines` to locs_sf
index <- st_nearest_feature(locs_sf, highway$osm_lines)

# Use that index to ask for the distance to that object
locs_sf$road_dist_m <- st_distance(locs_sf, highway$osm_lines[index,], 
                                   by_element=T) # Note `by_element=T` tells st_distance to evaluate things line by line. 
```

#### water bodies

We also might want to calculate the distances to the nearest water body, and important resource for wildlife. We can do that using the following:

```{r ch4_32, class.source="Rmain"}
water <- opq(aoi) %>%
           add_osm_feature(key = 'natural', value = 'water') %>%
           osmdata_sf()

```

Lets check our data:

```{r ch4_33, class.source="Rinfo"}
par(mfrow=c(1,2))
plot(st_geometry(water$osm_points), main="osm_points")
#plot(st_geometry(water$osm_lines), main="osm_lines") # lines is empty and wont plot
plot(st_geometry(water$osm_polygons), main="osm_polygons")
```

```{r, include=F}
# Reset plot params
par(mfrow=c(1,1))
```

In this instance, the lines is empty, out best bet are the points or the polygons options!

```{r ch4_34, class.source="Rmain"}
index <- st_nearest_feature(locs_sf, water$osm_polygons)

locs_sf$water_dist_m <- st_distance(locs_sf, water$osm_polygons[index,], by_element=T) # Note `by_element=T` tells st_distance to evaluate things line by line. 
```

For more examples of using the `osmdata` package see: [the projects github page](https://github.com/ropensci/osmdata)

### Vegetation productivity

#### modisfast

`modisfast` is an R package designed for easy and fast downloads of [MODIS](https://lpdaac.usgs.gov/data/get-started-data/collection-overview/missions/modis-overview/) Land products, [VIIRS](https://lpdaac.usgs.gov/data/get-started-data/collection-overview/missions/s-npp-nasa-viirs-overview/) Land products, and [GPM](https://gpm.nasa.gov/data) (Global Precipitation Measurement Mission) Earth Observation data. It allows for easy access to time series data accross large spatial extents directly to your computer! These are the data layers commonly used to extract normalized difference vegetation index (NDVI) and Enhanced Vegetation Index (EVI) information. When using `modisfast` you should reference:

[Taconet et al., (2024). modisfast: An R package for fast and efficient access to MODIS, VIIRS and GPM Earth Observation data. Journal of Open Source Software, 9(103), 7343,](https://doi.org/10.21105/joss.07343s)

For a quick into to the package see: <https://github.com/ptaconet/modisfast>

**IMPORTANT** to use this package you should create your own login for the Earthdata servers, then specify your EOSDIS credentials.

**To create your free account go to : <https://urs.earthdata.nasa.gov/>.**

Let's load the package:

```{r ch4_46, class.source="Rmain", eval=F}
library(modisfast)

# Specify your own credentials:
log <- mf_login(credentials = c("your_username", "your_password")) # set your own EOSDIS username and password

```

```{r ch4_46, class.source="Rmain", eval=T, include=F}
# Specify your own credentials:
log <- mf_login(credentials = c("cameratrap_course", "CMI_Revelstoke_2025")) # set your own EOSDIS username and password
```

For `modisfat` to work, we need to provide a sf polygon with an "id" column. We can do this easily using the `aoi` object we created earlier"

```{r ch4_47, class.source="Rmain"}
# Create aoi if sf format
aoi_sf <- st_as_sf(data.frame(id = "algar", geom = st_as_sfc(aoi), wkt = "geom", crs = 4326)) # a ROI of interest, format sf polygon

# MODIS collections and variables (bands) of interest
collection <- "MOD13Q1.061" # run mf_list_collections() for an exhaustive list of collections available
variables <- c("_250m_16_days_NDVI") # run mf_list_variables("MOD13Q1.061") for an exhaustive list of variables available for the collection "MOD13A3.061"

# Specify the time range - we will use a winter-summer transition
time_range <- as.Date(c("2019-03-01", "2019-07-01")) # a time range of interest

```

```{r}
## Get the URLs of the data
urls <- mf_get_url(
  collection = collection,
  variables = variables,
  roi = aoi_sf,
  time_range = time_range
)

## Download the data. By default the data is downloaded in a temporary directory, but you can specify a folder
res_dl <- mf_download_data(urls, parallel = TRUE)

# Import it into R
r <- mf_import_data(
  path = dirname(res_dl$destfile[1]),
  collection = collection,
  proj_epsg = 26912
)

# Plot the resulting NDVI data
terra::plot(r, col = rev(terrain.colors(20)), 
            range=c(0,1)) # Specify the range of values you want to plot

# Extract the data to camera locations
tmp <- terra::extract(r, locs_terra, fun=mean, ID=F) # Extract - each layer will become its own column
locs_sf$mean_ndvi <- rowMeans(tmp) # take the mean of each column

```

Easy!

Check the output:

```{r ch4_54, class.source="Rinfo"}
boxplot(locs_sf$mean_ndvi,
        ylab="Mean NDVI score",
        las=1)
```

Here we have created an average NDVI score across five months, but we could get monthly averages. Some products are even available at daily resolution!

### Digging deeper

If you want to dig into estimating NDVI metrics from camera trap viewshed, rather than from satellite data, check out the `phenopix` R package. It allows the user to extract visual information from time lapse images. It provides a quantitative daily measure of vegetation phenology at each site (e.g. green-up, senescence, snow cover).

[Alberton, B. et al. 2017. Introducing digital cameras to monitor plant phenology in the tropics: applications for conservation. Perspect. Ecol. Conserv](https://www.sciencedirect.com/science/article/pii/S2530064417300019)

[Filippa, G. et al. 2017. phenopix: Process Digital Images of a Vegetation Cover. R package version 2.3.1.](https://rdrr.io/rforge/phenopix/)

The Phenopix package has a five step process:

1.  a region of interest (ROI) is identified;
2.  the red, green, and blue digital numbers from each image in the time series is extracted and an index of relative ‘greenness’ is computed and plotted from the digital numbers;
3.  the vegetation indices’ data points are filtered to remove inconsistencies;
4.  a curve is fit to the data and phenophases are determined from the curve;
5.  phenophase uncertainties are calculated.

To see an application and comparison of these metrics, we highly recommend that you check out Catherine Sun's (WildCo alumni) paper on the subject:

[Sun, Catherine, et al. "Simultaneous monitoring of vegetation dynamics and wildlife activity with camera traps to assess habitat change." Remote Sensing in Ecology and Conservation 7.4 (2021): 666-684.](https://zslpublications.onlinelibrary.wiley.com/doi/10.1002/rse2.222)

And the code associated with this publication on the [WildCo GitHub Page](https://github.com/WildCoLab/WildCo_Phenology)

## Convert and save your covariates

```{r ch4_57, message=F, warning=F, class.source="Rmain"}
# Convert columns to numeric
locs_sf$road_dist_m <- as.numeric(locs_sf$road_dist_m)

# Convert it back to a dataframe
locs_sf$geometry <- NULL

locs <- left_join(locs, locs_sf)


# Write the dataset

write.csv(locs, paste0("data/processed_data/", locs$project_id[1],"_camera_locations_and_covariates.csv"), row.names=F)

```

## Correlations between predictors

So we have used a variety of different techniques to generate covariates for our subsequent analyses. However, it is important to note that we cannot just through these variables into a model.

One way to check if your different variables are confound/correlated is using the `corrplot` package.

```{r ch4_58, class.source="Rmain"}
library(corrplot)

# First we need to create a correlation matrix between the different variables of interest
M <- cor(locs[, c("line_of_sight_m", "water_depth_m", "elevation",
                     "road_dist_m", "mean_ndvi")])
```

Now lets make the basic `corrplot`:

```{r ch4_59, class.source="Rinfo"}
corrplot(M)
```

The cells denote pairwise correlations between the rows and the columns. The great thing about `corrplot` is customization option are near endless - see [the corrplot vignette](https://cran.r-project.org/web/packages/corrplot/vignettes/corrplot-intro.html).

Let's make a better, more informative, `corrplot`!

```{r ch4_60, class.source="Rmain"}

corrplot(M,                              #The correlation matrix we made
         method="color",                 # How we want the cells 
         type="upper",                   # Just show the upper part (it is usually mirrored)
         order="hclust",                 # Order the variables using the hclust method
         addCoef.col = "black",          # Add coefficient of correlation  
         tl.col="black", tl.srt=45,      # Control the text label color and rotation
         diag=F                          # Suppress the diagonal correlations (which are 1 anyway)
         )

```

In general there is very low correlation between our different predictors! If we were seeing pairwise correlations \>0.7 we perhaps wouldn't include those in the same model.
