---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Habitat use

```{r, echo=F, results='hide', message =F, warning=F}
# Check you have them and load them
list.of.packages <- c("kableExtra", "tidyr", "ggplot2", "gridExtra", "lme4", "dplyr", "Hmsc")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
lapply(list.of.packages, require, character.only = TRUE)

```

Camera traps are well suited for the quantification of habitat use accross multiple species. To assess habitat use, we typically quantify the detection rate, or a relative abundance index (RAI). RAI estimates are fairly simple to estimate, and understand, than other metrics (such as occupancy)and thus their use is widespread. 

In its simpliest form this represents the number of indpendent events (or individuals within indepnendent events) of a given species at a given camera, divided by the number of days that camera was active during that period of interest. 

Such abundance indices are typically analysed in a linear modelling framework, for single species, increasingly, for multispecies assemblages. Addressed in turn below. 

We will use the "processed_data/Algar_30min_Independent_Monthly_counts.csv" dataframe as the starting point for this data analysis. Which looks like this:

```{r}
monCount <- read.csv("data/processed_data/Algar_30min_Independent_Monthly_counts.csv", header=T)
```

Which, as a quick reminder, looks like this:

```{r, echo=F}
kbl(head(monCount))%>%
  kable_paper() %>%
  scroll_box(height = "200px")
```

So our rows represent a month at each given site, and we have the number of camera days each location was trapped. This is also the format which most liner model analysis packages require you to have your data in. Easy!

Our next step to create the relative abundance index. We will divide each count by the number of days the station was active in each month then multiply by the a standardised number of days - often people use 100.

In R this would look like:

```{r}
# Create a dataframe to store these 'relative abundances'
monRAI <- monCount
# Divide the species abundances (which start in column four), by the amount of camera effort
monRAI[ , 4:ncol(monCount)] <- (monRAI[ , 4:ncol(monCount)]/monRAI$Effort)*100
```

We can then examine the relationship between raw counts (on the x-axis) with our relative abundance index (on the y-axis), using *Odocoileus virginianus* as an example:

```{r}
plot(monRAI$Odocoileus.virginianus ~ monCount$Odocoileus.virginianus,
     las=1, pch=19, ylab="RAI", xlab="Raw")
abline(a=0, b=3.22, lty=2, col="grey")
     
```

So in months where we have an incomplete sample (say 20 days in January instead of 31), the RAI will be higher than the dashed grey line.

## Single species habitat use model
The most common way to analyse habitat use data is through linear models. There are a variety if differnt frameworks to fit and compare different linear models to address a host of different hypotheses, but if you are just starting out you should be aware of two key packages:

- `lme4` -> frequentist and information theoretic approaches
- `brms` -> bayseian approaches

There is no right or wrong about which package and which approach you use to test your hypotheses, just make sure you understand the implications of your choices!

**RESOURCES To Go HERE**

**Example in lme4**

In this worked example we will analyse how mammal capture vary using a mixed effects model. We will not test the model assumptions or interograte the findings, there are better resources to allow you to do that! Here we simply demonstrate how to use our camera datat to fit the model.

First we must install the packages we require: 'lme4' and `tidyr':

```{r}
 library(lme4); library(tidyr)
```

The lme4 package requires a dataframe format (as above), with the response term and the predictor variables all included in the same location. In this example we will explore if the habitat use of *Odocoileus virginianus* varied based on treatment strata.

**Preparing our data**

Recall that the information about each camera is recorded in the Station_covariares file:
```{r}
sta <- read.csv("data/raw_data/Example_station_data.csv", header=T)

kbl(head(sta))%>%
  kable_paper() %>%
  scroll_box(width = "750px", height = "200px")

```

So we have three variables to pick from:

- `Treatment` is the feature where the camera trap was deployed: HumanUse = a camera on a seismic line used and maintained in an "open" state by humans; Offline = a camera in contiguous forest >200m from a seismic line 	
- `LOW500` is the proportion of surrounding habitat 500m around the camera which is classified as "Lowland" 
- `LOS` is the average straight line distance in m which you can view from the camera trap location.  

In this example we will start by keeping things simple and just exploring if `LOW500` influences the habitat use of *Odocoileus virginianus*.

Our first task is to link the information in the `sta` dataframe, with that in the monCount dataframe. To do this we will use `left_join()` in the dplyr package - using the camera stations (`Deployment.Location.ID` column) to link the data. 

```{r}
library(dplyr)
modDat <- left_join(monCount, sta[c("Deployment.Location.ID", "LOW500")])
```

Lets check it has worked, the `LOW500` column should be added on the right hand side:

```{r}
kbl(head(modDat))%>%
  kable_paper() %>%
  scroll_box(width = "750px", height = "200px")
```


Next we will fit a mixed effects model to this dataset using `lme4`. You may have noticed that we havent used the monRAI dataframe we fit earlier! That is because we can create a relative abundance index within the model itself by providing an `offset()` - which functions in an equivalent way to what we did beforehand, and preseveres the original units of the obsevations (counts). 

The model takes the form:
  
  Response term ~ fixed effect + offset + (1|random intercept), data frame, distribution

We include `Deployment.Locaion.ID` as the random intercept, as camera stations are repeatedly sampled at monthly intervals and thus our data (rows in the dataframe) are not independent. We use the `poisson`  family, as our response term is a count.  

```{r}
m1 <- glmer(Odocoileus.virginianus ~ LOW500 + offset(log(Effort)) + (1|Deployment.Location.ID) , data=modDat, family="poisson")
```

We can view a summary of the model fit using:

```{r}
summary(m1)
```

As stated at the start of this guide, we are not focusing on whether the models we apply are appropriate or finding "the best" models for this datatset, so do not spend too much time trying to interpret this infomation! 

We can plot the relationship between by keeping things simple and just exploring if `LOW500` influences the habitat use of *Odocoileus virginianus* habitat using the predict function and `ggplot2`.


```{r}
newDat <- cbind(expand.grid(LOW500=seq(min(sta$LOW500),max(sta$LOW500), length.out=50)),Effort=100)

# Type "response" gives predictions on the original (count) scale.
newDat$Pred <- predict(m1,newdata=newDat,re.form=NA,
                  type="response")

plot(newDat$Pred~newDat$LOW500, type="l",
     ylim=c(0, max(newDat$Pred)), lwd=2,
     las=1, ylab="Predicted habitat use",
     xlab="Proportion lowland habitat")

```

So our model suggests that the habitat use of *Odocoileus virginianus* decreases as the proportion of loaland habitat increases. This is consistent with our understanding of White-tailed deer, which prefer upland habitats!

If you want more applied examples of generating predictions from mixed effects models, check out [Ben Bolkers workbook](https://bbolker.github.io/mixedmodels-misc/ecostats_chap.html) 

```{r, echo=F, eval=F}

# Maybe add in the future: 

#Bootstrapped confidence intervals from the Bolker link


# Bootstrap some CI's
#set.seed(101)
#m_bb <- bootMer(m1,
#              FUN=function(x)
#              predict(x,re.form=NA,newdata=newDat,
#              type="response"),
#              nsim=400)
#
#m_CI<- t(apply(g_bb$t,2,quantile,c(0.025,0.975),na.rm=TRUE))
#newDat <- cbind(newDat, m_CI)
```

## Further things to consider

- bayesian models - brms materials
- model fit
- overdispersion
- standardising input variables


## Example multispecies model

Given advances in computer power and analytic approaches, it is becoming increasingly popular to model multiple species within the same frameowrk. This opens up a variety of things not previously possible. 

As with single species linear models, there are many choices available for modeeling multiple species in the same frameowrk. Two notable options are:

- GJAM
- HMSc

In this example we will use the `Hmsc` package.

```{r}
library(Hmsc)
```

**Preparing our data**

The format of data required for joint species distribution models is different to that of traditional linear models. 

Rather than lumping the response term and fixed effects within the the same dataframe, we need a separate `Y` matrix of site.time x species, and an `Xdata` dataframe of your fixed and random effects variables. 

```{r}
# Pull the count data into its own matrix
Y <- as.matrix(monCount[, 4:ncol(monCount)])
# You can make the rown names the Site.Date values 
# (just incase you want to check things)
row.names(Y) <- paste(monCount$Deployment.Location.ID, monCount$Date, sep=".")
```

Which looks like this:

```{r}
kbl(head(Y))%>%
  kable_paper() %>%
  scroll_box(width = "750px", height = "200px")
```

We then create the XData in a similar way to before, but this time dropping the species information:

```{r}
Xdat <- left_join(monCount[c("Deployment.Location.ID", "Date", "Effort")], sta[c("Deployment.Location.ID", "LOW500")])

```

Which looks like:

```{r}
kbl(head(Xdat))%>%
  kable_paper() %>%
  scroll_box(width = "750px", height = "200px")
```


With Bayesian approaches we need to set up our sampling conditions

```{r}
nChains   = 2 
thin      = 2 
samples   = 100 
transient = 10*thin
verbose   = T
```

Setup our random effect:

```{r}
# Add a station-level random effect (for the covariances)
studyDesign = data.frame(station = as.factor(Xdat$Deployment.Location.ID))
rL = HmscRandomLevel(units = studyDesign$station)
```

Specify our model""

```{r}
# Model specification
str(Xdat)

m = Hmsc(Y = Y, XData = Xdat[,c("LOW500", "Effort")], 
         XFormula = ~LOW500 + log(Effort),
         studyDesign = studyDesign, ranLevels = list(station = rL), 
         distr="poisson")


m = sampleMcmc(m, thin = thin, samples = samples, transient = transient,
               nChains = nChains, verbose = verbose)

```


```{r}
postBeta = getPostEstimate(m, parName = "Beta")
par(mar=c(8,12,1,1))
plotBeta(m, post = postBeta, param = "Support", supportLevel = 0)
```








## Other resources

Palmer, Meredith S., et al. "Evaluating relative abundance indices for terrestrial herbivores from large‐scale camera trap surveys." African journal of ecology 56.4 (2018): 791-803. 
