---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Habitat use

```{r, echo=F, results='hide', message =F, warning=F}
# Check you have them and load them
list.of.packages <- c("kableExtra", "tidyr", "ggplot2", "gridExtra", "lme4", "dplyr", "Hmsc")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
lapply(list.of.packages, require, character.only = TRUE)

```

Camera traps are well suited for the quantification of habitat use across multiple species. To assess habitat use, we typically quantify the detection rate - the number of dections divided by the time interval of interest. Detection rates are fairly simple to estimate and understand thus their use is widespread. 

In its simplest form this represents the number of independent events (or individuals within independent events) of a given species at a given camera, divided by the number of days that camera was active during that period of interest. 

Such detection rates are thought to reflect the habitat use of a species at a given location - extreme care should be taken if you want to use it as an indesx of abundance. Detection rates are typically analysed in a linear modelling framework for single species, and increasingly for multispecies too. 

We will use the "processed_data/Algar_30min_Independent_Monthly_counts.csv" dataframe as the starting point for this data analysis. Which looks like this:

```{r}
monCount <- read.csv("data/processed_data/Algar_30min_Independent_Monthly_counts.csv", header=T)
```

Which, as a quick reminder, looks like this:

```{r, echo=F}
kbl(head(monCount))%>%
  kable_paper() %>%
  scroll_box(height = "200px")
```

So our rows represent a month at each given site, and we have the number of camera days each location was trapped. This is also the format which most liner model analysis packages require you to have your data in. Easy!

Our next step to create the relative abundance index. We will divide each count by the number of days the station was active in each month then multiply by the a standardized number of days - often people use 100.

In R this would look like:

```{r}
# Create a dataframe to store these detection rates
monDR <- monCount
# Divide the species abundances (which start in column four), by the amount of camera effort
monDR[ , 4:ncol(monCount)] <- (monDR[ , 4:ncol(monCount)]/monDR$Effort)*100
```

We can then examine the relationship between raw counts (on the x-axis) with our detection rate (on the y-axis), using *Odocoileus virginianus* as an example:

```{r}
plot(monDR$Odocoileus.virginianus ~ monCount$Odocoileus.virginianus,
     las=1, pch=19, ylab="DR", xlab="Raw")
abline(a=0, b=3.22, lty=2, col="grey")
     
```

So in months where we have an incomplete sample (say 20 days in January instead of 31), the DR will be higher than the dashed grey line.

## Single species habitat use model
The most common way to analyse habitat use data is through linear models. There are a variety if different frameworks to fit and compare different linear models to address a host of different hypotheses, but if you are just starting out you should be aware of two widely used packages:

- `lme4` -> frequentest and information theoretic approaches 
- `brms` -> Bayesian approaches

There is no right or wrong about which package and which approach you use to test your hypotheses, just make sure you understand the implications of your choices!

**Example in lme4**

In this worked example we will analyse how habitat use varies using a mixed effects model. We will not test the model assumptions or interrogate the findings, there are better resources to allow you to do that! Here we simply demonstrate how to use our camera data to fit the model.

First we must install the packages we require: 'lme4' and `tidyr':

```{r}
 library(lme4); library(tidyr)
```

The lme4 package requires a dataframe format (as above), with the response term and the predictor variables all included in the same location. In this example we will explore if the habitat use of *Odocoileus virginianus* varied based on treatment strata.

**Preparing our data**

Recall that the information about each camera is recorded in the Station_covariates file:
```{r}
sta <- read.csv("data/raw_data/Example_station_data.csv", header=T)

kbl(head(sta))%>%
  kable_paper() %>%
  scroll_box(width = "750px", height = "200px")

```

So we have three variables to pick from:

- `Treatment` a catagorical variable which reflects strata where the camera trap was deployed: HumanUse = a camera on a seismic line used and maintained in an "open" state by humans; Offline = a camera in contiguous forest >200m from a seismic line 	
- `LOW500` is a continuous variable reflecting the proportion of surrounding habitat 500m around the camera which is classified as "Lowland" 
- `LOS` is a continuous variable reflecting the average stDRght line distance (in metres) which you can view from the camera trap location.  

In this example we will start by keeping things simple and just exploring if `LOW500` (the proportion of lowland habitat) influences the habitat use of *Odocoileus virginianus*. We expect that it would, as White-tailed deer prefer upland habitat types in this landscape.

Our first task is to link the information in the `sta` dataframe, with that in the monCount dataframe. To do this we will use `left_join()` in the dplyr package - using the camera stations (`Deployment.Location.ID` column) to link the data. 

```{r}
library(dplyr)
modDat <- left_join(monCount, sta[c("Deployment.Location.ID", "LOW500")])
```

Lets check it has worked, the `LOW500` column should be added on the right hand side:

```{r}
kbl(head(modDat))%>%
  kable_paper() %>%
  scroll_box( height = "200px")
```


Next we will fit a mixed effects model to this data set using `lme4`. You may have noticed that we haven't used the monDR dataframe we made earlier! That is because we can create a relative abundance index within the model itself by providing an `offset()` term . An offset functions to scale the response term based on the amount of survey effort, and perseveres the original units of the observations (counts). 

The model takes the form:
  
  Response term ~ fixed effect + offset() + (1|random intercept), data frame, distribution

We include `Deployment.Locaion.ID` as the random intercept, as camera stations are repeatedly sampled at monthly intervals and thus our data (rows in the dataframe) are not independent. We use the `poisson`  family, as our response term is a count.  

```{r}
m1 <- glmer(Odocoileus.virginianus ~ LOW500 + offset(log(Effort)) + (1|Deployment.Location.ID) , data=modDat, family="poisson")
```

We can view a summary of the model fit using:

```{r}
summary(m1)
```

As stated at the start of this guide, we are not focusing on whether the models we apply are appropriate or finding "the best" models for this datasheet, so do not spend too much time trying to interpret this information! 

We can plot the relationship between by keeping things simple and just exploring if `LOW500` influences the habitat use of *Odocoileus virginianus* habitat using the predict function and `ggplot2`.

To do this we create a dummy dataframe of "new" data (`newDat`):

```{r}
newDat <- cbind(expand.grid(LOW500=seq(min(sta$LOW500),max(sta$LOW500), length.out=50)),Effort=100)

# Type "response" gives predictions on the original (count) scale.
newDat$Pred <- predict(m1,newdata=newDat,re.form=NA,
                  type="response")

plot(newDat$Pred~newDat$LOW500, type="l",
     ylim=c(0, max(newDat$Pred)), lwd=2,
     las=1, ylab="Predicted habitat use",
     xlab="Proportion lowland habitat")

```

So our model suggests that the habitat use of *Odocoileus virginianus* decreases as the proportion of lowland habitat increases. This is consistent with our understanding of White-tailed deer, which prefer upland habitats!

If you want more applied examples of generating predictions from mixed effects models, check out [Ben Bolkers workbook](https://bbolker.github.io/mixedmodels-misc/ecostats_chap.html) 

```{r, echo=F, eval=F}

# Maybe add in the future: 

#Bootstrapped confidence intervals from the Bolker link


# Bootstrap some CI's
#set.seed(101)
#m_bb <- bootMer(m1,
#              FUN=function(x)
#              predict(x,re.form=NA,newdata=newDat,
#              type="response"),
#              nsim=400)
#
#m_CI<- t(apply(g_bb$t,2,quantile,c(0.025,0.975),na.rm=TRUE))
#newDat <- cbind(newDat, m_CI)
```

### Future additions

- Bayesian models - brms materials
- model fit
- over dispersion`
- standardizing input variables
- continuous and catagorical examples
- Spatial and temporal non-independence 
- Confidence/credible intervals

### Further reading


## Example multispecies model

Given advances in computer power and analytic approaches, it is becoming increasingly popular to model multiple species within the same framework. This opens up a variety of things not previously possible. 

As with single species linear models, there are many choices available for modeling multiple species in the same framework. Two notable options are:

- [GJAM](https://cran.r-project.org/web/packages/gjam/vignettes/gjamVignette.html) 
- [HMSc](https://www2.helsinki.fi/en/researchgroups/statistical-ecology/hmsc)

In this example we will use the `Hmsc` package.

```{r}
library(Hmsc)
```

**Preparing our data**

The format of data required for joint species distribution models is different to that of traditional linear models. 

Rather than lumping the response term and fixed effects within the the same dataframe, we need a separate `Y` matrix of site.time x species, and an `Xdata` dataframe of your fixed and random effects variables. 

```{r}
# Pull the count data into its own matrix
Y <- as.matrix(monCount[, 4:ncol(monCount)])
# You can make the rown names the Site.Date values 
# (just incase you want to check things)
row.names(Y) <- paste(monCount$Deployment.Location.ID, monCount$Date, sep=".")
```

Which looks like this:

```{r}
kbl(Y)%>%
  kable_paper() %>%
  scroll_box(height = "200px")
```

We then create the XData in a similar way to before, but this time dropping the species information:

```{r}
Xdat <- left_join(monCount[c("Deployment.Location.ID", "Date", "Effort")], sta[c("Deployment.Location.ID", "LOW500")])
```

Which looks like:

```{r}
kbl(Xdat)%>%
  kable_paper() %>%
  scroll_box(height = "200px")
```


With Bayesian approaches we need to set up our sampling conditions

```{r}
nChains   = 2 
thin      = 2 
samples   = 100 
transient = 10*thin
verbose   = T
```

Setup our random effect:

```{r}
# Add a station-level random effect (for the covariances)
studyDesign = data.frame(station = as.factor(Xdat$Deployment.Location.ID))
rL = HmscRandomLevel(units = studyDesign$station)
```

Specify our model""

```{r, message =F, warning=F, eval=F}
# Model specification
m = Hmsc(Y = Y, XData = Xdat[,c("LOW500", "Effort")], 
         XFormula = ~LOW500 + log(Effort),
         studyDesign = studyDesign, ranLevels = list(station = rL), 
         distr="poisson")


m = sampleMcmc(m, thin = thin, samples = samples, transient = transient,
               nChains = nChains, verbose = verbose)

```


```{r, echo=F, message=F, warning=F}
# Model specification
m = Hmsc(Y = Y, XData = Xdat[,c("LOW500", "Effort")], 
         XFormula = ~LOW500 + log(Effort),
         studyDesign = studyDesign, ranLevels = list(station = rL), 
         distr="poisson")

m = sampleMcmc(m, thin = thin, samples = samples, transient = transient,
               nChains = nChains, verbose = F)

```


We can plot a basic summary of the modelled effects using the following code.

```{r}
postBeta = getPostEstimate(m, parName = "Beta")
par(mar=c(8,12,1,1))
plotBeta(m, post = postBeta, param = "Support", supportLevel = 0)
```

We the colours denote the size and magnitude of the effect of proportion of lowland habitat. *NOTE* treat these results with cauthion as the number of model runs is very low (to increase speed) and the model assumptions have not been interrograted. 

## Further reading

Palmer, Meredith S., et al. "Evaluating relative abundance indices for terrestrial herbivores from large‐scale camera trap surveys." African journal of ecology 56.4 (2018): 791-803. 
