---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Occupancy

```{r, echo=F, results='hide', message =F, warning=F}
# Check you have them and load them
list.of.packages <- c("kableExtra", "tidyr", "ggplot2", "gridExtra", "lme4", "dplyr", "unmarked", "lubridate")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
lapply(list.of.packages, require, character.only = TRUE)

```

Name a better duo than that camera trapping and occupancy modelling. I'll wait.

Occupancy modelling has been the mainstay of analysing data from camera traps since 

**Assumptions**

Ideally camera stations would be indpendent of one another, you would should only capture a given individual on a single camera. Thus for single species studies research space cameras by a single "home range" - although this in itself is tricky to quantify.  

ETC

## Single species occupancy model

In this example we will use the weekly observations dataframe fro the SingleSiteExploration script.

```{r}
weekObs <- read.csv("data/processed_data/Algar_30min_Independent_Weekly_observations.csv", header=T)
```

Which, as a quick reminder, looks like this:

```{r}
kbl(weekObs)%>%
  kable_paper() %>%
  scroll_box( height = "200px")
```

We first need to create a site x occasion matrix for our focal species, *Odocoileus virginianus*, using a 7 day occasion length. This break our camera data into seven day periods. We can visualise what this entailes using the camera activity plots in the SingleSiteExploration script:

```{r, echo=F}
# Import the original effort data
eff <- read.csv("data/raw_data/Example_deployment_data.csv", header=T)
tz<- "UTC"
# Put the time is POSIX format
eff$Camera.Deployment.Begin.Date <- as.Date(ymd_hms(eff$Camera.Deployment.Begin.Date, truncated=3, tz=tz))
eff$Camera.Deployment.End.Date   <- as.Date(ymd_hms(eff$Camera.Deployment.End.Date, truncated=3, tz=tz))

# Count the number of camera ststions
n.stat <- length(unique(eff$Deployment.Location.ID))

# Plot them

# Adjust layout
par(mar=c(6,6,1,1))
plot(c(min(eff$Camera.Deployment.Begin.Date, na.rm=T), 
       max(eff$Camera.Deployment.End.Date, na.rm=T)), 
       c(1,n.stat), las=1, ylab="", xlab="", type="n", yaxt="n", xaxt="n")

# Put them in alphabetical order
plot.order <- unique(eff$Deployment.Location.ID)
plot.order <- rev(plot.order[order(plot.order)])

axis(2, at= 1:n.stat, labels= plot.order, las=1, cex.axis=0.8)

# Make nice x-axis labels
tmp <- as.Date(paste0(substr(c(min(eff$Camera.Deployment.Begin.Date, na.rm=T), 
       max(eff$Camera.Deployment.End.Date, na.rm=T)),1,7),"-01"))

axis(1, at= seq(tmp[1], tmp[2], by="month"), 
     labels= substr(seq(tmp[1], tmp[2], by="month"),1,7), las=2, cex.axis=0.8)


#Horizontal lines for months
abline(v=as.numeric(seq(min(eff$Camera.Deployment.Begin.Date, na.rm=T),max(eff$Camera.Deployment.End.Date, na.rm=T), by=7)), col=rgb(0,0,0,0.1))


#mtext("Camera Deployment ID", 2, 4)
# Make lines for each of the cameras
for(i in 1:length(plot.order))
{
  abline(h=i, col=rgb(0,0,0,0.1))
  tmp <- eff[eff$Deployment.Location.ID==plot.order[i],]
  for(j in 1:nrow(tmp))
    {
      lines(c(tmp$Camera.Deployment.Begin.Date[j],
                       tmp$Camera.Deployment.End.Date[j]),
            c(i,i), lwd=2)
    }
  
}


```
Where each think black line is an active camera, and each thin grey box represents each slice of the detection history. 

We can create the detection histories using the following code:
```{r, message=F, warning=F}
library(dplyr)
library(tidyr)
library(magrittr)
library(tibble)

##Subset to your focal data
tmp <- weekObs[,c("Deployment.Location.ID", "Date", "Odocoileus.virginianus")]

# Turn the counts into presence/absence
tmp$Odocoileus.virginianus[tmp$Odocoileus.virginianus>0] <- 1

# Create a history
detHist <-  tmp %>%
  #Convert from long into wide format (a detection history) using spread
  spread(Date,Odocoileus.virginianus, fill = NA) %>% 
  # group by deployment location
  group_by(Deployment.Location.ID,) %>%
  column_to_rownames( var = "Deployment.Location.ID")

detHist <- as.matrix(detHist)
```

The resulting data frame looks like this:

```{r, echo=F}
kbl(detHist)%>%
  kable_paper() %>%
  scroll_box( height = "200px")
```

It is a matrix of all the dates the cameras were active, and whether the species was present or absent on that day. It should mirror the camera activity plot above. The `fill = NA` command puts a zero where there is data for a given day. However, we have lost our effort information - the number of days each camera was active in a given time period.

To keep that information we also create an effort history `effHist`:

```{r}
# To create the effort matrix - inst of the Focal Species bring in the effort
effDat <- weekObs[,c("Deployment.Location.ID", "Date", "Effort")]

effHist <-  effDat %>%
  # Create a matrix based on dates and effort
  spread(Date,Effort, fill = NA) %>% 
  # group by deloyment Location ID, then make that the row.namesd
  group_by(Deployment.Location.ID,) %>%
  column_to_rownames( var = "Deployment.Location.ID") 

effHist <- as.matrix(effHist)
```

Check that it looks sensible:

```{r, echo=F}
kbl(effHist)%>%
  kable_paper() %>%
  scroll_box( height = "200px")
```


Remove weeks w/0 full sample?
```{r}
detHist[effHist!=7] <- NA

```

### unmarked package

```{r}
library(unmarked)
sta <- read.csv("data/raw_data/Example_station_data.csv", header=T)
# Unmarked wants your detection history, effort data and site covariates as matrices.

# Build an unmarkedFramOccu
unDat <- unmarkedFrameOccu(y = detHist,
                          # siteCovs = dataframe with site rows x column variables (in same order as detHist)
                           siteCovs = sta) 
```



```{r}
# Fit general model all variables
m1 <- occu(formula = ~1 # detection formula first
                     ~LOW500, # occupancy formula second,
                data = unDat)
```

```{r}
summary(m1)

```

```{r}
# Generate new data to predict from 
newDat <- cbind(expand.grid(LOW500=seq(min(sta$LOW500),max(sta$LOW500), length.out=100)))

newDat <- predict(m1, type="state", newdata = newDat, appendData=TRUE)

p1 <- ggplot(newDat, aes(x = LOW500, y = Predicted)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.5, linetype = "dashed") +
  geom_path(size = 1) +
  labs(x = "Proportion lowland habitat", y = "Occupancy probability") +
  theme_classic() +
  coord_cartesian(ylim = c(0,1))

p1
```


## Multispecies occupancy model

Coming soon!
