# On your own

Congratulations! You made it to the end of the planned content in one peice.

You now have the opportunity to apply some of the skills you have learnt to a completely new database. Cut and paste your code from previous chapters to check the data, write the analysis date frames and explore the dataset.

I will help you read in the data - and correct a mistake I made. Then you are on your own!

## The new dataset


```{r ch13_1, class.source="Rmain"}
# Load your data 
pro <- read.csv("data/raw_data/your_data/proj.csv", header=T)
pro$project_id <- pro$project_name

img <- read.csv("data/raw_data/your_data/img.csv", header=T)
img$project_id <- pro$project_name

dep <- read.csv("data/raw_data/your_data/dep.csv", header=T)
dep$project_id <- pro$project_name

cam <- read.csv("data/raw_data/your_data/cam.csv", header=T)
cam$project_id <- pro$project_name

```

Load your packages:

```{r ch13_2, class.source="Rmain", message=F, warning=F, results='hide'}
# A list of the required packages
list.of.packages <- c("activity", "corrplot", "cowplot", "dplyr",  "elevatr", "gfcanalysis",   "ggplot2", "gridExtra", "iNEXT", "kableExtra", "Hmsc", "leaflet", "lme4", "lubridate", "magrittr", "MCMCvis", "MODISTools", "osmdata", "pals", "plotly", "remotes", "rmarkdown", "sf", "spOccupancy", "stars", "stringr", "terra", "tibble", "tidyr",  "unmarked", "viridis", "jtools", "vegan", "MuMIn")

# A check to see which ones you have and which are missing
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]

# Code which tells R to install the missing packages
if(length(new.packages)) install.packages(new.packages)
lapply(list.of.packages, require, character.only = TRUE)

```

Away you go!

```{r, echo=F, eval=F}
# start dates
dep$start_date <- ymd(dep$start_date)

# end dates
dep$end_date   <- ymd(dep$end_date)

dep$days <- interval(dep$start_date, dep$end_date)/ddays(1)


img$timestamp <- ymd_hms(img$timestamp)


# First, set a single categorical variable of interest from station covariates for summary graphs. If you do not have an appropriate category use "project_id".
category <- "feature_type"

# We first convert this category to a factor with discrete levels
dep[,category] <- factor(dep[,category])
# then use the turbo() function to assign each level a color
col.cat <- turbo(length(levels(dep[,category])))
# then we apply it to the dataframe
dep$colours <- col.cat[dep[,category]]

m <- leaflet() %>%
  addProviderTiles(providers$Esri.WorldImagery, group="Satellite") %>%  
  addTiles(group="Base") %>%     # Include a basemap option too
  addCircleMarkers(lng=dep$longitude, lat=dep$latitude,
                   # Co lour the markers depending on the 'feature type'
                   color=dep$colours,
                   # Add a popup of the placename and feature_type together 
                   popup=paste(dep$placename, dep[,category])) %>%
  
  # Add a legend explaining what is going on
  addLegend("bottomleft", colors = col.cat,  labels = levels(dep[,category]),
                   title = category,
                   labFormat = labelFormat(prefix = "$"),
                   opacity = 1) %>%
  
  # add a layer control box to toggle between the layers
  addLayersControl(
                    baseGroups = c("Satellite", "Base"))

m


# Distance between traps
camera_locs <- dep %>% 
  dplyr::select(placename, latitude, longitude) %>% 
  unique() %>% # remove duplicated rows (rows where the placename and coordinates match)
  st_as_sf(coords = c("longitude", "latitude"), crs = "+proj=longlat") # Convert to `sf` format

camera_dist <- st_distance(camera_locs) %>% 
                  as.dist() %>% 
                  usedist::dist_setNames(as.character(camera_locs$placename)) %>% 
                  as.matrix()

# convert to pairwise list
camera_dist_list <- t(combn(colnames(camera_dist), 2))
camera_dist_list <- data.frame(camera_dist_list, dist = camera_dist[camera_dist_list]) %>% 
                          arrange(dist) # sort descending

# keep just the closest camera for each location
camera_dist_list <- camera_dist_list %>% 
                  group_by(X1) %>% 
                  slice(which.min(dist))


summary(camera_dist_list$dist)


table(unique(img$placename) %in% unique(dep$placename))

table(unique(dep$placename)  %in% unique(img$placename))
dep[!dep$placename %in% img$placename,]

# One station - I will have to fix!


# ACtivity check


# Call the plot
p <- plot_ly()

# We want a separate row for each 'placename' - so lets turn it into a factor
dep$placename <- as.factor(dep$placename)

# loop through each place name
for(i in seq_along(levels(dep$placename)))
  {
      #Subset the data to just that placename
      tmp <- dep[dep$placename==levels(dep$placename)[i],]
      # Order by date
      tmp <- tmp[order(tmp$start_date),]
      # Loop through each deployment at that placename
      for(j in 1:nrow(tmp))
      {
        # Add a line to 'p'
        p <- add_trace(p, 
                       #Use the start and end date as x coordinates
                       x = c(tmp$start_date[j], tmp$end_date[j]), 
                       #Use the counter for the y coordinates
                       y = c(i,i), 
                       # State the type of chart
                       type="scatter",
                       # make a line that also has points
                       mode = "lines+markers", 
                       # Add the deployment ID as hover text
                       hovertext=tmp$deployment_id[j], 
                       # Color it all black
                       color=I("black"), 
                       # Suppress the legend
                       showlegend = FALSE)
      }
      
  }
# Add a categorical y axis
 p <- p %>%   layout(yaxis = list(

      ticktext = as.list(levels(dep$placename)), 

      tickvals = as.list(1:length(levels(dep$placename))),

      tickmode = "array"))


p



### DETCTION CHECK



#Camera activity plot
# Make a separate plot for each 20 stations For each 20 stations
# To do this make a plot dataframe
tmp <- data.frame("deployment_id"=unique(dep$deployment_id), "plot_group"=ceiling(1:length(unique(dep$deployment_id))/20))

dep_tmp <- left_join(dep,tmp, by="deployment_id")

for(i in 1:max(dep_tmp$plot_group))
{  
  # Call the plot
  p <- plot_ly() 
  
  #Subset the data to just that placename
  tmp <- dep_tmp[dep_tmp$plot_group==i,]
  # Order by placename 
  tmp <- tmp[order(tmp$placename),]
  
 
 # Loop through each deployment at that placename
  for(j in 1:nrow(tmp))
    {
        #Subset the image data
        tmp_img <- img[img$deployment_id==tmp$deployment_id[j],]
        
        if(nrow(tmp_img)>0)
        {
         
          p <- add_trace(p, 
                       #Use the start and end date as x coordinates
                       x = c(tmp_img$timestamp), 
                       #Use the counter for the y coordinates
                       y = rep(j, nrow(tmp_img)), 
                       # State the type of chart
                       type="scatter",
                       # make a line that also has points
                       mode = "markers", 
                       # Add the deployment ID as hover text
                       hovertext=paste(tmp_img$genus,tmp_img$species), 
                       # Color it all black
                       marker = list(color = "red"), 
                       # Suppress the legend
                       showlegend = FALSE)
        }
        
       # Add a line to 'p'
        p <- add_trace(p, 
                       #Use the start and end date as x coordinates
                       x = c(tmp$start_date[j], tmp$end_date[j]), 
                       #Use the counter for the y coordinates
                       y = c(j,j), 
                       # State the type of chart
                       type="scatter",
                       # make a line that also has points
                       mode = "lines", 
                       # Add the deployment ID as hover text
                       hovertext=tmp$deployment_id[j], 
                       # Color it all black
                       color=I("black"), 
                       # Suppress the legend
                       showlegend = FALSE)
      }
  # Add custom y axis labels  
  p <- p %>%   layout(yaxis = list(

      ticktext = as.list(tmp$deployment_id), 

      tickvals = as.list(1:nrow(tmp)),

      tickmode = "array"))
  
  print(p)
      
  
} 

##### Already got common names dont worry about that


# First lets convert our timestamp to decimal hours
img$hours <- hour(img$timestamp) + minute(img$timestamp)/60 + second(img$timestamp)/(60*60)

# Count all of the captures
tmp <- img %>% group_by(common_name) %>% summarize(count=n())

yform <- list(categoryorder = "array",
              categoryarray = tmp$common_name)

fig <- plot_ly(x = img$hours, y = img$common_name,type="scatter",
               height=1000, text=img$deployment_id, hoverinfo='text',
               mode   = 'markers',
               marker = list(size = 5,
                             color = 'rgba(50, 100, 255, .2)',
                             line = list(color = 'rgba(0, 0, 0, 0)',
                                         width = 0))) %>% 
              layout(yaxis = yform)
fig

# Remove the column
img$hours <- NULL





```




